BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2025-05-01 manuscript/BarnebyLives -COPY2 ACTUALLY CLEAN.rmd
script to clean up the problem records identified in manuscript to accession

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract: |
  \noindent
  **Premise:** Depositing specimens in herbaria is a time-consuming process. Many institutions have reduced the amount of funding for herbaria and universities have reduced the amount of education dedicated to collections and curatorial tasks. However, the continual accessioning of herbarium specimens is essential for current and future research on ecology, evolution, and conservation. To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the ease of accessioning specimens for collectors.  
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. The Census Bureau, the United States Geological Survey, and other sources to supplement collection notes by providing information on administrative jurisdictions, abiotic site conditions and locality information. It uses in-house spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclature innovations. Optionally, the package writes driving directions to the sites using Google Maps. The package outputs data in tabular and spatial formats for review by the user before rendering labels using LaTeX.        
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users with current taxonomic opinions, generates driving directions to sites, and produces aesthetically pleasing labels and shipping manifests in a matter of minutes.       
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
  - \usepackage{endfloat}
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{rotating}
---

@@

Nearly 400 million specimens are housed worldwide in herbaria [@thiers2021herbaria]. 
However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity and towards understanding cellular and molecular processes underpinning life [@prather2004decline; @pyke2010biological; @daru2018widespread]. 
This shift, among other factors, led to a decline in the funding allocated to collection-based research, the number of staff maintaining and accessing new collections, and educating students in these practices [@funk2014erosion].
Historically, specimens have been used to describe the taxonomic diversity of plants and document global floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
However, renewed interest in herbarium collections utilizing 'big data approaches,’ such as museuomics, has brought herbaria back to the forefront of the natural sciences and grearly expanded their roles in science [@ronsted2020integrative; @marsico2020small].

Innovations in specimen digitization, data sharing, computing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
The current use of specimens and their ancillary data extends well beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium]. 
We suspect that collections are yet to realize their full potential, and as currently novel approaches, such as electronic and remote sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid]. 
While image-based or purely observational (rather than collection-based) citizen science approaches (e.g., iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs, specimens contain rich data that are not accessible via images. 
Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen. 
These factors will ensure that the specimens remain the premier botanical data source into perpetuity.

However, despite renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline].
We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via commercially developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
While many young botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observe that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
Some of the apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].

The generation of an herbarium specimen involves many steps that are easy to take for granted [@forman1989herbarium].
For example, while acquiring appropriate political information for a collection site appears simple, young collectors rarely have adequate cartographic resources (printed topographic maps or GIS software) at their disposal.
In topographically complex areas, where administrative borders are often associated with hydrological basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report administrative details in error. 
Even finding appropriate site names can rarely be resolved without a printed map, as many navigation-related software now consider most features that would serve as site names extraneous. 
Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied scientific names, even when these names are unanimously accepted by taxonomic specialists in the group and other regional curators [@hitchcock2018flora]. 
Furthermore, formatting a label correctly (e.g., author abbreviations, italicization, etc.) is a time-consuming process with many opportunities to introduce errors in formatting which reduce the apparent credibility of a collector. 
Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, for example, applying italicization. 
Even if a collector successfully navigates all these hurdles, the time allocated to each step is quite large, and may discourage them from further collecting.

As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases and to speed up the generation of labels.
BarnebyLives rapidly provides political and administrative boundary information for a collection site using data from the U.S. Census Bureau [@walker2022tigris], the Public Land Survey System (PLSS), and ownership details of public lands via the Protected-Areas Database (PAD-US) [@usgs2024padus].
Site names are suggested by finding the closest unambiguously named place feature in the Geographic Name Information System (GNIS) and the precise calculation of distance and azimuth from this feature to the collection site [@gnis2024].
Using the Global Mountain Biodiversity Assessment (GMBA) Mountain Inventory V. 2, a standardized named mountain data set with global coverage allows for a relevant descriptor of the general region with less ambiguity [@snethlage2022hierarchical].
Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the resolved species may be searched via Kew's Plant of the World Online for relevant synonyms [@govaerts2021world; @powo2024]. 
Author abbreviations are verified using the International Plant Names Index (IPNI) Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviations of authorities [@ipni2024; @powo2024].
Checks to search for and flag common issues associated with spreadsheet software or data transcription, such as the auto-filling of coordinate and date columns. 
After a final review of the data, flagged or generated by the package, it allows for the option to export spreadsheets that are suitable for mass uploading of data to multiple common herbarium databases as well as the generation of herbarium labels.

Currently, to our knowledge label generation functionality is provided explicitly by two programs, PLabel and Symbiota, and by the Microsoft Word tool Mail Merge [@perkins2020Plabel; @gries2014symbiota]. 
The office suite costs money, and in our experience, is finicky; further, its functionality ends with label creation. 
PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality. 
However, beyond verifying the nations of collection it does not include data cleaning functionalities.  
While some sources indicate that it can only be used on Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems. 
Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries’ (see Figure 1). 
However, not all herbaria use Symbiota and many have original database systems that they maintain (for example, Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html; Missouri Botanical Garden https://tropicos.org/specimen/Search; and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/). 
However, and most importantly many collectors prefer to generate their own labels, especially as they are likely to send different sets of collections to different institutions. 
Accordingly, the functionality of Symbiota should exist in an ecosystem with alternative systems.  In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats after utilizing BLs data cleaning utilities.

@@

## Usage
All steps of BarnebyLives, except for label generation are run within the freely available RStudio. 
Data may be read from any common spreadsheet management system or database connection such as Excel, or free alternatives such as LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github Pages (https://sagesteppe.github.io/BarnebyLives/) and around 100 real-world examples are on a Google Sheets accessible from the page. 
BarnebyLives is atypical for R packages in that it requires a considerable amount of data to operate (Table 1). 
Virtually all on-disk memory associated with the package are used to store spatial data.
The amount of spatial data varies according to the domain that the user decides to support (Figure 3).
Functions that require on-disk data require a path to data as an argument.  Manually supplying the path argument allows users to determine an appropriate storage location suitable for their needs.

We anticipate that for a typical user, BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 3-arc second (~90m) resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives. 
Given that the attributes which the package collects data on are tailored to the Western U.S. region, we do not expect local installs to exceed the size of ours. 
The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
Unfortunately, we have not found Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system, but suspect Ubuntu subsystem for Windows may allow for integration of these tools.

@@

The spatial module has five required functions and two optional functions.  
*autofill_checker* searches for patterns in the input latitude and longitude data associated with autofilling from various spreadsheet programs and will emit a warning if they are encountered.  
*coords2sf* creates a spatially explicit simple feature (sf) geometry dataset for the input data. 
*political_grabber* determines many levels of administrative ownership, including land management and public land survey system sections.  
*physical_grabber* provides various geographic data, such as elevation, landform position, and aspect using 90m resolution spatial data.  
*site_writer*  write distance and azimuth to collection site from the nearest official named place from the GNIS database. 
*directions_grabber* is an optional function that writes driving directions from a reasonably sized town to the closest drivable area to the site using the Google Maps API, which will require a valid Google account that is free per month for most personal and smaller academic usages.  
*dms2dd* is an optional function used to convert from coordinates denoted in the degrees minutes and second format (for example, 42°08'39.9\"N 87°47'08.3\"W) to decimal degree format (for example 42.14439, -87.78569).

Please note that the function *physical_grabber* is the one portion of the package where a decoupling may exist between the collection site, and the resolution of the spatial data. 
While we expect the mismatch to be negligible for all effective purposes relating to: elevation, major geology type, and in general aspect, estimates of slope at this resolution may be biased - generally to lower angles.
For these reasons collectors must always make notes on the truly local environment which taxa are found in, and consider that the notes from BL reflect the greater landscape which a microfeature may be present in. 
While this mis-match will seldom effect landscape ecologists, it may have implications for other data users.

@@

The formatting module has three functions. Two are optional; however, they are run locally and so quickly that there is no reason to skip them. 
*date_parser* parses an input date into various formats for notating collection and determination dates on labels.
*associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks. 
*field_lengths* will emit messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.

@@

The data exporting module contains three functions that interact with LaTeX templates and require slightly more advanced R user interactivity, such as setting up mapping functions using the tidyverses purrr package.  
*labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.  
*label_writer* write from a flatfile or spreadsheet to small 4x4 inch herbarium labels (users can modify these dimensions as they see fit). 
*format_database_import* will write out a spreadsheet of cleaned data in a variety of formats, currently: Jepson, Symbiota, and Consortium of Pacific Northwest herbaria are supported.

@@

subs_batch <- c(1586, 1880, 2187, 2159)

data <- read.csv('data/test_data.csv', na.strings = "") %>% 
  drop_na(c('Longitude', 'Latitude', 'Date_digital')) %>% 
  unite(col = 'Scientific_name', 
        c(Binomial, Infrarank, Infraspecies), na.rm=TRUE, sep = " ", remove = F) %>% 
  filter(Scientific_name != '')#, Collection_number %in% subs_batch)#, Collection_number %in% subs_batch)

n_families <- data %>% 
  group_by(Family) %>% 
  count() %>% 
  nrow() # 74 families

n_genera <- data %>% 
  separate(Scientific_name, into = c('Genus', 'Epithet') , remove = FALSE, extra = 'drop') %>% 
  group_by(Genus) %>% 
  count() %>% 
  nrow() # 292 Genera

n_spp <- data %>% 
  group_by(Binomial) %>% 
  count() %>% 
  nrow() # 616 species

n_infraspecies <- data %>% 
  drop_na(Infraspecies) %>% 
  group_by(Binomial, Infraspecies) %>% 
  count() %>% 
  nrow() # 66 distinct infraspecies

n_sp_authors <- data %>% 
  drop_na(Binomial_authority) %>% 
  count() %>% # 557 groups of authors
  pull()

n_infra_sp_auths <- data %>% 
  drop_na(Infraspecific_authority) %>% 
  group_by(Binomial, Infraspecies) %>% 
  count() %>% 
  nrow() # 22 distinct infra species author groups

n_vegetation_fields <- data %>% 
  drop_na(Vegetation) %>% 
  distinct(Vegetation) |>
  nrow()

n_associates_fields <- data %>% 
  drop_na(Associates) %>% 
  distinct(Associates) |>
  nrow()

# number of collection sites

n_sites <- data %>% 
  distinct(Latitude, Longitude) %>% 
  nrow()

@@

![The spatial extent-or domain- (orange), and herbarium collection sites (burgundy) tested in this manuscript.](../graphics/plots/collections_map.png){width=50%}

The testing of the package within this manuscript was performed using a subset of the authors collections from 2018-2022, while most development was performed on their 2023 and 2024 collections. 
Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used for most functionality. 
In total `r nrow(data)` records were used for testing various functions, these records were from `r n_sites` sites located across Western North America (Figure 2). 
In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families.

@@

saveRDS(time_powo_searcher, file = 'data/processed/time_powo_searcher')
saveRDS(data, file = 'data/processed/data_w_POWO_search')

time_directions_grabber <- system.time({ # write directions to sites
  SoS_gkey = Sys.getenv("Sos_gkey")
  data <- directions_grabber(data, api_key = SoS_gkey)
})

saveRDS(time_directions_grabber, file = 'data/processed/time_directions_grabber')
saveRDS(data, file = 'data/processed/data_w_Google_Maps')

@@

# also save a spreadsheet of the cleaned data, we'll use this to create the labels. 
write.csv(data, file = 'data/processed/cleaned_data.csv', row.names = FALSE)
```

```{r Total Time of operations, eval = F}

time_powo_searcher <- readRDS(file = 'data/processed/time_powo_searcher')
time_directions_grabber <- readRDS('data/processed/time_directions_grabber')

time_trials <- data.frame(as.matrix(do.call(rbind, mget(ls(pattern = '^time'))))) %>% 
  rownames_to_column('Function') %>% 
  select(-user.child, -sys.child) %>% 
  mutate(Function = str_remove(Function, 'time_')) %>% 
  mutate(Module = case_when(
    Function %in% c('associate_dropper', 'split_binomials', 'date_parser') ~ 'Style',
    Function %in% c('autofill_checker', 'coords2sf', 'dms2dd', 'physical_grabber', 'time_geodata_writer',
                    'political_grabber', 'site_writer', 'geodata_writer', 'directions_grabber') ~ 'Geospatial', 
    Function %in% c('spell_check', 'family_spell_check', 'author_check',  'powo_searcher', 
                    'associates_spell_check_veg', 'associates_spell_check_ass') ~ 'Taxonomic',
    Function %in% c('time_label_maker', 'time_shipping_manifest', 'time_label_assembly') ~ 'Labels'
    ),
    Group = if_else(Function %in% c('powo_searcher', 'directions_grabber'), 'Online', 'Local')
  )

# save these as RDS so operations do not need to be run each time document knits
saveRDS(time_trials, file = 'data/processed/time_trials')

rm(time_trials)
rm(list=ls(pattern = 'time'))
```

```{r Create Labels, eval = F}
processed <- read.csv('data/processed/cleaned_data.csv') %>% 
  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number), 
         Coordinate_Uncertainty = '+/- 5m') %>% 
  filter(Collection_number != 'sine non') %>% 
  data.frame()

@@

# now copy bash script to location # note we are one dir up from the lowest level. 
# note that I customized the label template after import, so we will not run this again. 
# p2libs <- system.file(package = 'BarnebyLives')
# folds <- file.path( 'rmarkdown', 'render_labels.sh')

# file.copy(from = file.path(p2libs, folds), 
#           to = '/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw')
```

```{r combine labels, eval = F}
## this SHOULD be a sh chunk, so launch in a shell! I just want to be sure you can see the code. 
## and eval = TRUE will be respected. 
user: pwd
/media/steppe/hdd/Barneby_Lives-dev/manuscript/labels/raw
# this will have both your script and a dir full of raw labels with a users name
# now launch the bash script with:

# or chmod it etc!
# et voila! data will be saved up on dir in a 'final' directory. 
# we timed it like this:
\time -o ../process2final.txt  bash render_labels.sh collector='Reed'
```

@@

While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate [@patten2024geographic].
We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to assume that responsibility in a relatively fast and streamlined format.
By utilizing both R and LaTeX and having publicly available source code on Github, this program allows users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications in project branches.

>>>>

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract: |
  \noindent
  **Premise:** Depositing specimens in herbaria is a time-consuming process. Many institutions have reduced the amount of funding for herbaria and universities have reduced the amount of education dedicated to collections and curatorial tasks. However, the continual generation of herbarium specimens is essential for current and future research on ecology, evolution, and conservation. To facilitate the continued growth of herbaria, BarnebyLives was developed as a tool to supplement collection notes, perform geographic and taxonomic information processing, enact spell checks and other QC steps, produce labels, and submit digital data to increase the ease of accessioning specimens.  
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. The Census Bureau for administrative jurisdictional data, and other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions and locality information. It uses in-house spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclature innovations. Optionally, the package writes driving directions to the sites using Google Maps. The package outputs data in tabular and spatial formats for review by the user before rendering labels using LaTeX.        
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users with current taxonomic opinions, generates driving directions to sites, and produces aesthetically pleasing labels and shipping manifests in a matter of minutes.       
keywords: |
  herbarium, natural history museum, collections, geospatial, R, LaTeX   
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
  - \usepackage{endfloat}
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{rotating}
---

@@

Nearly 400 million specimens are housed worldwide in herbaria [@thiers2021herbaria]. 
However, The rate of accessioning new collections to herbaria diminished in the 20^th^ century as priorities in biology shifted away from describing and documenting earths biodiversity and towards understanding cellular and molecular processes underpinning life [@prather2004decline; @pyke2010biological; @daru2018widespread]. 
This shift, among other factors, led to a decline in the funding allocated to collection-based research, the number of staff maintaining and accessing new collections, and educating students in these practices [@funk2014erosion].
Historically, specimens have been used to describe the taxonomic diversity of plants and document global floristic diversity [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative]. 
However, renewed interest in herbarium collections utilizing 'big data approaches,’ such as museuomics, has brought herbaria back to the forefront of the natural sciences [@ronsted2020integrative; @marsico2020small].

Innovations in specimen digitization, data sharing, computing, DNA sequencing, and statistics have perhaps brought about greater use of herbarium specimens than ever before [@greve2016realising; @james2018herbarium; @brewer2019factors; @ronsted2020integrative].
The current use of specimens and their ancillary data extends well beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand the ecological niches, phenological processes, and interactions of plants [@ronsted2020integrative; @davis2023herbarium]. 
We suspect that collections are yet to realize their full potential, and as currently novel approaches, such as remote and electronic sensing and meta-barcoding, become more accessible the use of collections will increase [@tosa2021rapid]. 
While image-based or purely observational (rather than collection-based) citizen science approaches (e.g., iNaturalist, BudBurst) have recently dovetailed with herbarium specimens to meet many current research needs, specimens contain rich data that are not accessible via images. 
Only specimens have the ability to: provide samples of DNA, secondary metabolites, or proteins, material for measuring (micro-)morphological attributes [@borges2020schrodinger], and seeds or pollen.  These factors ensure that the specimens remain the premier botanical data source.

However, despite renewed recognition of the utility of collections, efforts to grow them appear slow [@prather2004decline].
We conjecture that this is partly because collecting and depositing specimens is a fundamentally slower process, especially for novice collectors, relative to taking photographs via professionally developed apps on smartphones [@daru2018widespread; @mishler2020spatial; @manzano2021fair].
While many young botanists are capable of using dichotomous keys and other resources to reliably identify and collect satisfactory  material, we observed that they face difficulties navigating several aspects of data acquisition, processing, and preparation of labels for submission to herbaria.
Evidently, apparent problems include the lack of dedicated time at the end of a field season to process specimens, a general lack of education on cartography and orienteering, natural history (e.g., geology, geomorphology), nomenclature, and familiarity with various computer programs (for example, Microsoft Office suite), and increasing foundational knowledge of plant systematics and phylogenetics [@nanglu2023nature; @woodland2007botanists; @barrows2016crossroads].

The generation of an herbarium specimen involves many steps that are easy to take for granted [@forman1989herbarium].
For example, while acquiring appropriate political information for a collection site appears simple, young collectors rarely have adequate cartographic resources (printed topographic maps or GIS software) at their disposal.
In topographically complex areas, where administrative borders are often associated with hydrological basins and the ridges defining them, collectors are liable to misinterpret their true geographic position and report details in error. 
Even finding appropriate site names can rarely be solved without a printed map, as many navigation-related software now consider many features that would serve as site names extraneous. 
Similarly, the rate at which taxonomic innovations occur, the volume of the literature, and the reluctance of some regional curators to embrace a phylogenetic approach to plant classification have made it difficult to find more recently applied names, even when these names are unanimously accepted by taxonomic specialists in the group [@hitchcock2018flora]. 
Furthermore, formatting a label correctly (e.g., author abbreviations) is a time-consuming process with many opportunities to introduce errors in formatting. 
Anecdotally, many mail merge templates offered by herbaria still require collectors to modify many variables by hand, for example, applying italicization. Even if a collector navigates all these hurdles, the time allocated to each step is quite large.

As a result of these concerns, we have developed an R package, BarnebyLives, that aims to increase both the quality of data rendered to labels and recorded in databases and to speed up the process of producing labels.
It rapidly provides political and administrative boundary information for a collection site using data from the U.S. Census Bureau [@walker2022tigris], the Public Land Survey System (PLSS), and ownership details of public lands via the Protected-Areas Database (PAD-US) [@usgs2024padus].
Site names are suggested by finding the closest unambiguously named place feature in the Geographic Name Information System (GNIS) and the precise calculation of distance and azimuth from this feature to the collection site [@gnis2024].
Using the Global Mountain Biodiversity Assessment (GMBA) Mountain Inventory V. 2, a standardized named mountain data set with global coverage allows for a relevant descriptor of the general region with less ambiguity [@snethlage2022hierarchical].
Spell checks on all scientific names (including associated species) are performed using a copy of the World Checklist of Vascular Plants, and the resolved species may be searched via Kew's Plant of the World Online for relevant synonyms [@govaerts2021world; @powo2024]. 
Author abbreviations are verified using the International Plant Names Index (IPNI) Standard Author Abbreviation Checklist and also returned by Kew's Plants of the World Online to ensure proper abbreviations of authorities [@ipni2024; @powo2024].
Checks to search for and flag common issues associated with spreadsheet software or data transcription, such as the auto-filling of coordinate and date columns. 
After a final review of the data, flagged or generated by the package, it allows for the option to export spreadsheets that are suitable for mass uploading of data to multiple common herbarium databases as well as the generation of herbarium labels.

Currently, label generation functionality is provided explicitly by two programs, PLabel and Symbiota, as well as by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota). 
The office suite costs money, and in our experience, is finicky; further, its functionality ends with label creation. 
PLabel is a standalone program that has greatly enhanced functionality relative to a mail merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface (GUI) functionality. 
However, it does not include data cleaning functionalities beyond verifying nations of collection. 
While some sources indicate that it can only be used in Microsoft, we expect it to be usable on Linux and Mac using Windows 'emulators' like Wine. 
The increasingly popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality in an attractive GUI web portal allowing for live management of collections and bypassing the need for a local installation, allowing it to be accessed on all operating systems. 
Symbiota offers functionality similar to the first four of our five stages of our 'Taxonomic' module and to our knowledge a check of the 'Political Boundaries’ (see Figure 1). 
However, not all herbaria use Symbiota and many have original database systems that they maintain (for example, Harvard University Herbarium, https://kiki.huh.harvard.edu/databases/specimen_index.html; Missouri Botanical Garden https://tropicos.org/specimen/Search; and The Consortium of Pacific Northwest Herbaria https://www.pnwherbaria.org/). 
However, and most importantly many collectors prefer to generate their own labels, especially as they are likely to send different sets of collections to different institutions. 
Accordingly, the functionality of Symbiota should exist in an ecosystem with alternative systems.  In scenarios where users want to keep rendering labels in either of the three existing alternatives, they can easily export data in the appropriate formats after utilizing BLs data cleaning utilities.

@@

## Usage
All steps of BarnebyLives, except for label generation are run within the freely available RStudio. 
Data may be read from any common spreadsheet management system or database connection such as Excel, or free alternatives such as LibreOffice, OpenOffice, or via the cloud on Google Sheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on a Github page (https://github.com/sagesteppe/BarnebyLives 'Input Data Column Names') and 96 real-world examples are on a Google Sheets accessible from the page. 
BarnebyLives is atypical for R packages in that it requires a considerable amount of data to operate (Table 1). 
Virtually all on-disk memory associated with the package are used to store spatial data.
The amount of spatial data varies according to the domain that the user decides to support (Figure 3).
Functions that require on-disk data require a path to data as an argument.  Manually supplying the path argument allows users to determine an appropriate storage location suitable for their needs.

We anticipate that for a typical user, BarnebyLives will require less than a couple gigabytes of memory (ours covering all of the conterminous Western U.S. at 3-arc second (~90m) resolution is ~16 GiB), while the processing requires relatively little RAM; hence, we believe installations can work on hardware as limited as Chromebooks, while having the data stored entirely on thumb-drives. 
The final steps of BarnebyLives, generating the labels, requires working installations of R Markdown, a LaTeX installation (e.g. [pdfTeX](https://www.tug.org/applications/pdftex/), [LuaTeX](https://www.luatex.org/), [XeLaTeX](https://www.overleaf.com/learn/latex/XeLaTeX)), and the open source command line tools [pdfjam](https://github.com/rrthomas/pdfjam) and [pdftk](https://linux.die.net/man/1/pdftk).
While these steps are run through a shell scripting language such as bash, we have wrapped them in R functions that bypass the need to enter the commands directly into a shell terminal outside of RStudio.
Unfortunately, we have not found Windows alternatives to pdfam and pdftek, so we are unable to offer the final label-generating functionality on that operating system.

@@

The spatial module has five required functions and two optional functions.

*coords2sf* creates a spatially explicit simple feature (sf) geometry dataset for the input data. political_grabber determines many levels of administrative ownership, including land management and public land survey system sections.  
*physical_grabber* provides various geographic data, such as elevation, landform position, and aspect using 90m resolution spatial data.  
*political_grabber*  grabs a wide range of administrative information for a collection site. 
*site_writer*  write distance and azimuth to collection site from the nearest official named place from the GNIS database. 
*directions_grabber* is an optional function that writes driving directions from a reasonably sized town to the closest drivable area to the site using the Google Maps API, which will require a valid Google account that is free per month for most personal and smaller academic usages. 
*dms2dd* is an optional function used to convert from coordinates denoted in the degrees minutes and second format (for example, 42°08'39.9\"N 87°47'08.3\"W) to decimal degree format (for example 42.14439, -87.78569). 
*autofill_checker* searches for patterns in the input latitude and longitude data associated with autofilling from various spreadsheet programs and will emit a warning if they are encountered.

@@

The formatting module has three functions. The first two we will detail are optional; however, they are run locally and so quickly that there is no reason to skip them. 
*associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks. 
*field_lengths* will emit messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.
*date_parser* is mandatory and parses an input date into various formats for notating collection and determination dates on labels.

@@

The data exporting module contains three functions that interact with LaTeX templates and require slightly more advanced R user interactivity, such as setting up mapping functions using the tidyverses purrr package.
*labels_skeleton* is an R 'script' which will require a few modification steps to tailor to each institution, these R scripts will put data into a user specified template, and serve as the interface to LaTeX.

@@

subs_batch <- c(1207, 2818, 2747, 2748, 2760:2770, 2782:2784, 2435, 2517, 1424, 
                2644, 2038, 2326, 2618)

data <- read.csv('data/test_data.csv', na.strings = "") %>% 
  drop_na(c('Longitude', 'Latitude', 'Date_digital')) %>% 
  unite(col = 'Scientific_name', 
        c(Binomial, Infrarank, Infraspecies), na.rm=TRUE, sep = " ", remove = F) %>% 
  filter(Scientific_name != '', Collection_number %in% subs_batch)#, Collection_number %in% subs_batch)

@@

![The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.](../graphics/plots/collections_map-trim.png){width=50%}

The package was released into beta testing using the primary authors collections from 2023. 
The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022. 
Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used. 
Resulting in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America (Figure 2). 
In total this data set had `r n_spp` species (with `r n_sp_authors` distinct sets of authors), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families which were used for testing.

@@

@@

# also save a spreadsheet of the cleaned data, we'll use this to create the labels. 
write.csv(data, file = 'data/processed/cleaned_data-problemchildren.csv', row.names = FALSE)
```

```{r Total Time of operations, eval = F}
rm(list=ls(pattern = 'time'))
```

```{r Create Labels, eval = F}
processed <- read.csv('data/processed/cleaned_data-problemchildren.csv.csv') %>% 
  mutate(UNIQUEID = paste0(Primary_Collector, Collection_number), 
         Coordinate_Uncertainty = '+/- 5m') %>% 
  filter(Collection_number != 'sine non') %>% 
  data.frame()

@@

```{r combine labels, eval = F}
# or chmod it etc!
# et voila! data will be saved up on dir in a 'final' directory. 
../process2final.txt  bash render_labels.sh collector='Reed'
```

@@

While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate (@patten2024geographic).
We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to do so in a relatively fast and streamlined format.
By utilizing both R and LaTeX and having publicly available source code on Github, this program allows users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications in project branches.
